\textbf{Proof Reconstruction in Classical Propositional Logic}

### 2017-04-30 slides/slides.tex
[ slides ] djvu fonts added.

\setmonofont[ExternalLocation=fonts/
  , BoldFont=SourceCodePro-Semibold.ttf
  , BoldItalicFont=SourceCodePro-SemiboldIt.ttf
  , ItalicFont=SourceCodePro-It.ttf
  ]{SourceCodePro-Semibold.ttf}
\newfontfamily\SourceCodeIt{fonts/SourceCodePro-It.ttf}
\newfontfamily\SourceCode{fonts/SourceCodePro-Regular.ttf}

@@

\title[Proof Reconstruction in Classical Propositional Logic]
  {\textbf{Proof Reconstruction in Classical Propositional Logic}
}
\subtitle{(Work in Progress)}
\date{Agda Implementors’ Meeting XXV\\
May 9-15th}
\author[Jonathan Prieto-Cubides and Andr\'es Sicard-Ram\'irez]{Jonathan Prieto-Cubides\\
Joint work with Andr\'es Sicard-Ram\'irez}
\institute{
Master in Applied Mathematics\\
Universidad EAFIT\\
Medell\'in, Colombia}
% --------------------------------------------------------------------------

@@

% \subsection{Proof Reconstruction}

% \subsubsection{Sledgehammer}
% \subsubsection{Waldemesiter}

% \subsection{Automatic Provers}

% \subsubsection{TPTP Syntax}

\begin{frame}{Bonus Slides}
\end{frame}

\begin{frame}[fragile, label=tptp-syntax]{TPTP Syntax}{Thousands of Problems for Theorem Provers}

@@

  \begin{itemize}
  \item Is a language\footnote{\url{http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html}}
        to encode problems \citep{Sut09}
  \item Is the input of the ATPs
  \item Annotated formulas with the form
   \begin{center}
\begin{tptp}
language(name, role, formula).
\end{tptp}
    \end{center}
    \begin{itemize}
      \item[\texttt{language}] FOF or CNF %THF, TFF,
      \item[\texttt{name}] to identify the formula within the problem
      \item[\texttt{role}] axiom, definition, hypothesis, conjecture, among others
      \item[\texttt{formula}] version in TPTTP format
    \end{itemize}
  \end{itemize}

@@

\begin{frame}[fragile, label=tptp-examples]{TPTP Examples}{From Prop-Pack repository
  \footnote{\url{http://github.com/jonaprieto/prop-pack}}}

@@

\begin{frame}[label=metis]{Metis Theorem Prover}{\url{http://www.gilith.com/software/metis/}}
 \tikz[overlay,remember picture]
  \node at (0.92\textwidth, 0.5cm)
    {\includegraphics[width=0.15\textwidth]{figures/metis}};

Metis is an automatic theorem prover for First-Order Logic with equality
\begin{block}{Why Metis?}
  \begin{itemize}
    \item \href{https://github.com/gilith/metis}{Open source} implemented in Standard ML
    \item Each refutation step is one of
      \href{http://www.gilith.com/software/metis/logical-kernel.txt}{\textbf{6 simple rules}}
    \item Reads problem in \hyperlink{tptp-syntax}{\texttt{TPTP} format}
    \item Outputs detailed proofs in \hyperlink{tstp-syntax}{\texttt{TSTP} format}
  \end{itemize}
\end{block}
\gobackreconstruction
\end{frame}

\begin{frame}[fragile, label=metis-rules]{Inference Rules of Metis}
\hyperlink{tstp-example}{TSTP} derivations by Metis exhibit these inferences
\footnote{Inference rules found in proofs of Propositional Logic theorems}:
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|cl| }
\hline
\textbf{Rule} & \textbf{Purpose} \\
\hline
\texttt{canonicalize} & transforms formulas to CNF, DNF or NNF\\
\texttt{clausify} & performs clausification\\
\hyperlink{atp-conjunct}{\texttt{conjunct}} & takes a formula from a conjunction\\
\texttt{negate} & applies negation to the formula\\
\texttt{resolve} & applies theorems of resolution\\
\texttt{simplify} & applies over a list of formula to simplify them\\
\texttt{strip} & splits a formula into subgoals\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:label}
\end{table}
\vfill
\gobackreconstruction
\end{frame}

\subsubsection{TSTP Derivations}

\begin{frame}[fragile, label=tstp-syntax]{TSTP Syntax}

@@

A TSTP derivation
\footnote{\url{http://www.cs.miami.edu/~tptp/TPTP/QuickGuide/Derivations.html}}
\begin{itemize}
  \item Is a \hyperlink{tstp-dag}{\textbf{D}irected \textbf{A}cyclic \textbf{G}raph} where
  \begin{itemize}
    \item[\texttt{leaf}] is a formula from the \hyperlink{tptp-syntax}{\texttt{TPTP}} input
    \item[\texttt{node}] is a formula inferred from parent formula
    \item[\texttt{root}] the final derived formula
  \end{itemize}
  \item Is a list of annotated formulas with the form:
  \end{itemize}

@@

where \textbf{\color{plum} source} typically is an inference record:
\begin{center}
{\footnotesize
\begin{tptp}
inference(rule, useful info, parents)
\end{tptp}
}
\end{center}
\gobackreconstruction
\end{frame}

@@

\begin{itemize}
  \item Proof found by \texttt{Metis} Prover for the problem $p ⊢ p$
{\small
\begin{tptp}
$ metis --show proof basic-4.tptp
fof(a, axiom, p).
fof(goal, conjecture, p).
fof(subgoal_0, plain, p),
  inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ p,
  inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, ~ p,
  inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p,
  inference(canonicalize, [], [a])).
fof(normalize_0_2, plain, $false,
  inference(simplify, [],
    [normalize_0_0, normalize_0_1])).
cnf(refute_0_0, plain, $false,
    inference(canonicalize, [], [normalize_0_2])).
\end{tptp}
}
\end{itemize}
\gobackreconstruction
\end{frame}

\begin{frame}[fragile, label=tstp-dag]{DAG for the previous TSTP derivation found by \texttt{Metis} Prover}

\vfill
By refutation, we proved $p ⊢ p$:
\begin{center}
\begin{prooftree}
\AxiomC{$p$}
\RightLabel{\texttt{assume}}
\UnaryInfC{$p$}
\RightLabel{\texttt{strip}}
\UnaryInfC{$p$}
\RightLabel{\texttt{negate}}
\UnaryInfC{$¬ p$}
\AxiomC{$p$}
\RightLabel{\texttt{canonicalize}}
\UnaryInfC{$p$}
\RightLabel{\texttt{simplify}}
\BinaryInfC{$⊥$}
\RightLabel{\texttt{canonicalize}}
\UnaryInfC{$⊥$}

@@

\begin{frame}[fragile, label=athena]{Athena}
  {\url{http://github.com/jonaprieto/athena}}

@@

Is a Haskell program
that translates proofs given by Metis Prover
in \texttt{TSTP} format to Agda code.\\

\begin{itemize}
    \item Parsing of \texttt{TSTP} language
    \item Creation and analysis of \hyperlink{tstp-dag}{\textbf{DAG}} derivations
    \item Analysis of inference rules used in the \texttt{TSTP} derivation
    \item Agda code generation
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|cl| }
\hline
\textbf{Library} & \textbf{Purpose} \\
\hline
\texttt{\hyperlink{agda-prop}{Agda-Prop}}& axioms and theorems of Classical Propositional Logic\\
\texttt{\hyperlink{agda-metis}{Agda-Metis}} & versions of the inference rules used by Metis\\
\hline
\end{tabular}}
}
\end{center}
\label{tab:library}
\end{table}

@@

\begin{frame}[fragile, label=agda-prop-2]{Agda-Prop Library}
  {\url{https://github.com/jonaprieto/agda-prop}}
\begin{itemize}
  \item Propositional Logic + PEM ($Γ ⊢ ϕ ∨ ¬ ϕ$)
  \item A data type for formulas
\begin{agda}
data Prop : Set where
  Var  : Fin n → Prop           -- Variables.
  $⊤$    : Prop                   -- Top (truth).
  $⊥$    : Prop                   -- Bottom (falsum).
  _$∧$_  : ($φ$ $ψ$ : Prop) → Prop    -- Conjunction.
  _$∨$_  : ($φ$ $ψ$ : Prop) → Prop    -- Disjunction.
  _$⇒$_  : ($φ$ $ψ$ : Prop) → Prop    -- Implication.
  _$⇔$_  : ($φ$ $ψ$ : Prop) → Prop    -- Biimplication.
  $¬$_   : ($φ$ : Prop) → Prop      -- Negation.
\end{agda}
\end{itemize}
\end{frame}

\begin{frame}[fragile, label=agda-prop]{Agda-Prop Library}
  {\url{https://github.com/jonaprieto/agda-prop}}
\begin{itemize}
  \item A data type for theorems
\begin{agda}
data _$⊢$_ : ($Γ$ : Ctxt)($φ$ : Prop) → Set
\end{agda}

\item Constructors
\begin{agda}
assume, axiom, weaken, $⊤$-intro, $⊥$-elim, $¬$-intro,
$¬$-elim, $∧$-intro, $∧$-proj$₁$, $∧$-proj$₂$, $∨$-intro$₁$,
$∨$-intro$₂$, $∨$-elim, $⇒$-intro, $⇒$-elim, $⇔$-intro,
$⇔$-elim$₁$, $⇔$-elim$₂$
\end{agda}
  \item Natural deduction proofs for more than 71 theorems
\begin{agda}
$⇔$-equiv, $⇔$-assoc, $⇔$-comm, $⇒-⇔-¬∨$, $⇔-¬$-to-$¬$,
$¬⇔$-to-$¬$, $¬¬$-equiv, $⇒⇒-⇔-∧⇒$, $⇔$-trans, $∧$-assoc,
$∧$-comm, $∧$-dist, $¬∧$-to-$¬∨¬$, $¬∨¬$-to-$¬∧$, $¬∨¬-⇔-¬∧$,
subst$⊢∧₁$, subst$⊢∧₂$, $∨$-assoc, $∨$-comm, $∨$-dist,
$∨$-equiv, $¬∨$-to-$¬∧¬$, $¬∧¬$-to-$¬∨$, $∨$-dmorgan,
$¬¬∨¬¬$-to-$∨$, cnf, nnf, dnf, RAA, ...
\end{agda}
\end{itemize}
\end{frame}

\begin{frame}[fragile, label=agda-metis]{Agda-Metis Library}
  {\url{https://github.com/jonaprieto/agda-metis}}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.7}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|clc|}
\hline
\textbf{Rule} & \textbf{Purpose} &\textbf{Theorem}\\
\hline
\texttt{canonicalize} & transforms formulas to CNF, DNF or NNF &\texttt{atp-canonicalize}\\
\texttt{clausify} & performs clausification &\texttt{atp-clausify}\\
\hyperlink{atp-conjunct}{\texttt{conjunct}} & takes a formula from a conjunction &\hyperlink{atp-conjunct}{\texttt{atp-conjunct}}\\
\texttt{negate} & applies negation to the formula &\texttt{atp-negate}\\
\texttt{resolve} & applies theorems of resolution &\texttt{atp-resolve}\\
\texttt{simplify} & applies over a list of formula to simplify them &\texttt{atp-simplify}\\
\texttt{strip} & splits a formula into subgoals &\texttt{strip}\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:agda-metis-table}
\end{table}
\vfill
\end{frame}

\begin{frame}[fragile, label=atp-conjunct]{Agda-Metis: Conjunct Inference}
   {\url{https://github.com/jonaprieto/agda-metis}}
\begin{itemize}
\item Definition
  \begin{equation*}
  conjunct (ϕ₁ ∧  ϕ₂ ∧ ⋯ ∧ \mathbf{ϕᵢ} ∧ ⋯ ∧ \phi_n, \mathbf{ϕᵢ}) ⟶ \mathbf{ϕᵢ}
  \end{equation*}

\item Function\footnote{\label{note-conjunct}Excerpt from the \texttt{Agda-Metis} library available in  \texttt{ATP.Metis.Rules.Conjunct} module.}:
\begin{agda}
conjunct : Prop → Prop → Prop
conjunct $(φ \wedge ψ)$ $ω$ with $⌊$ eq $φ$ $ω$ $⌋$ | $⌊$ eq ψ $ω$ $⌋$
... | true  | _     = $φ$
... | false | true  = $ψ$
... | false | false = conjunct $φ$ $ω$
conjunct $φ$ $ω$       = $φ$
\end{agda}
\item Theorem\textsuperscript{\ref{note-conjunct}}
\begin{agda}
atp-conjunct
  : ∀ $\{Γ\}$ $\{φ\}$
  → $(ω : Prop)$
  → $Γ \vdash φ$
  → $Γ \vdash$ conjunct $φ$ $ω$
\end{agda}
\end{itemize}
\end{frame}

@@

atp-conjunct $\{Γ\}$ $\{φ ∧ ψ\}$ $ω$ $Γ⊢φ$ with $⌊$ eq $φ$ $ω$ $⌋$ | $⌊$ eq $ψ$ $ω$ $⌋$
... | true  | _     = $∧$-proj$₁$ $Γ⊢φ$
... | false | true  = $∧$-proj$₂$ $Γ⊢φ$
... | false | false =
  atp-conjunct $\{Γ = Γ\}$ $\{φ = φ\}$ $ω$ ($∧$-proj$₁$ $Γ⊢φ$)
atp-conjunct {_} {Var x}    _ = id
atp-conjunct {_} {$⊤$}      _ = id
atp-conjunct {_} {$⊥$}      _ = id
atp-conjunct {_} {$φ ∨ φ₁$} _ = id
atp-conjunct {_} {$φ ⇒ φ₁$} _ = id
atp-conjunct {_} {$φ ⇔ φ₁$} _ = id
atp-conjunct {_} {$¬ φ$}    _ = id
\end{agda}
\vfill
\gobackreconstruction
\end{frame}

\begin{frame}[fragile, label=agda-file]{Agda Code Example}{Generated by Athena Tool}

\begin{itemize}
  \item The problem is $p ∧ q ⊢ q ∧ p$
  \item In TPTP format
\begin{tptp}
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
\end{tptp}
  \item A natural deduction proof
\begin{columns}
\begin{column}{0.8\textwidth}
 \begin{prooftree}
  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{$∧$-\texttt{proj}$₁$}
  \UnaryInfC{$ϕ$}

  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{$∧$-\texttt{proj}$₂$}
  \UnaryInfC{$ψ$}
  \RightLabel{$∧$-\texttt{intro}}
  \BinaryInfC{$ψ ∧ ϕ$}
  \end{prooftree}
\end{column}
\begin{column}{0.2\textwidth}  %%<--- here
   \begin{tikzpicture}[overlay, remember picture]
% \tikzset{shift={(current page.north)}, xshift=cm,yshift=-2cm}
\node[vertex, fill=black] (d1) {};
\node[vertex, fill=black, right=0.5cm of d1 ] (d2) {};
\node[vertex, fill=black, below=0.2cm of d1] (d3) {};
\node[vertex, fill=black, below=0.2cm of d2] (d4) {};
\node[vertex, fill=black, below right= 0.2cm and  0.25cm of d3] (d5) {};
\draw[-, thick] (d1) to (d3);
\draw[-, thick] (d2) to (d4);
\draw[-, thick] (d3) to (d5);
\draw[-, thick] (d4) to (d5);
\end{tikzpicture}
\end{column}
\end{columns}

@@

\begin{frame}[fragile, allowframebreaks]{TSTP derivation}
\begin{tptp}
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
fof(subgoal_0, plain, q,
    inference(strip, [], [goal])).
fof(subgoal_1, plain, q => p,
    inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ q,
inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, (~ q),
    inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p & q,
    inference(canonicalize, [], [a])).
fof(normalize_0_2, plain, q,
    inference(conjunct, [], [normalize_0_1])).
fof(normalize_0_3, plain, $false,
    inference(simplify, [],
        [normalize_0_0, normalize_0_2])).
cnf(refute_0_0, plain, $false,
    inference(canonicalize, [], [normalize_0_3])).
fof(negate_1_0, plain, ~ (q => p),
    inference(negate, [], [subgoal_1])).
fof(normalize_1_0, plain, ~ p & q,
    inference(canonicalize, [], [negate_1_0])).
fof(normalize_1_1, plain, p & q,
     inference(canonicalize, [], [a])).
fof(normalize_1_2, plain, p,
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_3, plain, q,
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_4, plain, $false,
    inference(simplify, [],
      [normalize_1_0, normalize_1_2, normalize_1_3])).
cnf(refute_1_0, plain, ($false),
    inference(canonicalize, [], [normalize_1_4])).
\end{tptp}
\end{frame}

>>>>

\setmonofont[ExternalLocation=fonts/
  , BoldFont=SourceCodePro-Semibold.ttf
  , BoldItalicFont=SourceCodePro-SemiboldIt.ttf
  , ItalicFont=SourceCodePro-It.ttf
  ]{SourceCodePro-Regular.ttf}
\newfontfamily\SourceCodeIt{fonts/SourceCodePro-It.ttf}
\newfontfamily\SourceCode{fonts/SourceCodePro-Regular.ttf}

@@

\title[Proof Reconstruction in Classical Propositional Logic]
  {\textbf{Proof Reconstruction in Classical Propositional Logic}
}
\subtitle{(Work in Progress)}
\date{Agda Implementors’ Meeting XXV\\
May 9-15th 2017}
\author[Jonathan Prieto-Cubides]{Jonathan Prieto-Cubides\\
Joint work with Andr\'es Sicard-Ram\'irez}
\institute{
Master in Applied Mathematics\\
Universidad EAFIT\\
Medell\'in, Colombia}
% --------------------------------------------------------------------------

@@

\begin{frame}[fragile, label=tptp-syntax]{\texttt{TPTP} Syntax}
  {Thousands of Problems for Theorem Provers}

@@

  \begin{itemize}
  \item Is a language\footnote{\url{http://www.cs.miami.edu/~tptp/TPTP/SyntaxBNF.html}}
        to encode problems \citep{Sut09}
  \item Is the input of the ATPs
  \item Annotated formulas with the form
   \begin{center}
\begin{tptp}
language(name, role, formula).
\end{tptp}
    \end{center}
    \begin{itemize}
      \item[\texttt{language}] FOF or CNF %THF, TFF,
      \item[\texttt{name}] to identify the formula within the problem
      \item[\texttt{role}] axiom, definition, hypothesis, conjecture
      \item[\texttt{formula}] version in \texttt{TPTP} format
    \end{itemize}
  \end{itemize}

@@

\begin{frame}[fragile, label=tptp-examples]{\texttt{TPTP} Examples}

@@

\begin{frame}[label=metis]{\texttt{Metis} Theorem Prover
  \footnote{\url{http://www.gilith.com/software/metis/}}}
 \tikz[overlay,remember picture]
  \node at (0.92\textwidth, 0.5cm)
    {\includegraphics[width=0.15\textwidth]{figures/metis}};

\texttt{Metis} is an automatic theorem prover for First-Order Logic with Equality
\citep{hurd2003first}
\begin{block}{Why Metis?}
  \begin{itemize}
    \item \href{https://github.com/gilith/metis}{Open source} implemented in
       \texttt{Standard ML}
    \item Each refutation step is one of
      \href{http://www.gilith.com/software/metis/logical-kernel.txt}{\textit{6 rules}}
    \item Reads problem in \hyperlink{tptp-syntax}{\texttt{TPTP} format}
    \item Outputs detailed proofs in \hyperlink{tstp-syntax}{\texttt{TSTP} format}
  \end{itemize}
\end{block}
\gobackreconstruction
\end{frame}

\begin{frame}[fragile, label=metis-rules]{Inference Rules of \texttt{Metis}}
\hyperlink{tstp-example}{TSTP} derivations by \texttt{Metis} exhibit these inferences
\footnote{Inference rules found in proofs of Propositional Logic theorems}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|ll| }
\hline
\textbf{Rule} & \textbf{Purpose} \\
\hline
\texttt{canonicalize} & transforms formulas to CNF, DNF or NNF\\
\texttt{clausify} & performs clausification\\
\hyperlink{atp-conjunct}{\texttt{conjunct}} & takes a formula from a conjunction\\
\texttt{negate} & applies negation to the formula\\
\texttt{resolve} & applies theorems of resolution\\
\texttt{simplify} & applies over a list of formula to simplify them\\
\texttt{strip} & splits a formula into subgoals\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:label}
\end{table}
\vfill
\gobackreconstruction
\end{frame}

\begin{frame}[fragile, label=tstp-syntax]{\texttt{TSTP} Syntax}

@@

A \texttt{TSTP} derivation
\footnote{\url{http://www.cs.miami.edu/~tptp/TPTP/QuickGuide/Derivations.html}}
\begin{itemize}
  \item Is a \hyperlink{tstp-dag}{\textbf{D}irected \textbf{A}cyclic \textbf{G}raph} where
  \begin{itemize}
    \item[\texttt{leaf}] is a formula from the \hyperlink{tptp-syntax}{\texttt{TPTP}} input
    \item[\texttt{node}] is a formula inferred from parent formula
    \item[\texttt{root}] the final derived formula
  \end{itemize}
  \item Is a list of annotated formulas with the form
  \end{itemize}

@@

where \texttt{\color{plum} source} typically is an inference record
\begin{center}
{\footnotesize
\begin{tptp}
inference(rule, useful info, parents)
\end{tptp}
}
\end{center}
\gobackreconstruction
\end{frame}

@@

\begin{itemize}
  \item Proof found by \texttt{Metis} Prover for the problem $p ⊢ p$
{\small
\begin{tptp}
$ metis --show proof problem.tptp
fof(a, axiom, p).
fof(goal, conjecture, p).
fof(subgoal_0, plain, p),
  inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ p,
  inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, ~ p,
  inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p,
  inference(canonicalize, [], [a])).
fof(normalize_0_2, plain, $false,
  inference(simplify, [],
    [normalize_0_0, normalize_0_1])).
cnf(refute_0_0, plain, $false,
    inference(canonicalize, [], [normalize_0_2])).
\end{tptp}
}
\end{itemize}
\gobackreconstruction
\end{frame}

\begin{frame}[fragile, label=tstp-dag]{DAG Example}

\vfill
By refutation, we proved $p ⊢ p$:
\begin{center}
\begin{prooftree}
\AxiomC{$p$}
\RightLabel{\texttt{\tiny assume}}
\UnaryInfC{$p$}
\RightLabel{\texttt{\tiny strip}}
\UnaryInfC{$p$}
\RightLabel{\texttt{\tiny negate}}
\UnaryInfC{$¬ p$}
\AxiomC{$p$}
\RightLabel{\texttt{\tiny canonicalize}}
\UnaryInfC{$p$}
\RightLabel{\texttt{\tiny simplify}}
\BinaryInfC{$⊥$}
\RightLabel{\texttt{\tiny canonicalize}}
\UnaryInfC{$⊥$}

@@

\begin{frame}[fragile, label=athena]{\texttt{Athena}
  \footnote{\url{http://github.com/jonaprieto/athena}}}

@@

Is a \texttt{Haskell} program
that translates proofs given by \texttt{Metis} Prover
in \texttt{TSTP} format to \texttt{Agda} code\\

\begin{itemize}
    \item Parsing of \texttt{TSTP} language
    \item Creation and analysis of \hyperlink{tstp-dag}{\textbf{DAG}} derivations
    \item Analysis of inference rules used in the \texttt{TSTP} derivation
    \item \texttt{Agda} code generation
\begin{table}[!ht]
\begin{center}
\scalebox{0.8}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|ll| }
\hline
\textbf{Library} & \textbf{Purpose} \\
\hline
\texttt{\hyperlink{agda-prop}{\texttt{Agda-Prop}}}& axioms and theorems of Classical Propositional Logic\\
\texttt{\hyperlink{agda-metis}{\texttt{Agda-Metis}}} & versions of the inference rules used by \texttt{Metis}\\
\hline
\end{tabular}}
}
\end{center}
\label{tab:library}
\end{table}

@@

\begin{frame}[fragile, label=agda-prop-2]{\texttt{Agda-Prop} Library
  \footnote{\url{https://github.com/jonaprieto/agda-prop}}}
\begin{itemize}
  \item Intuitionistic Propositional Logic + PEM ($Γ ⊢ ϕ ∨ ¬ ϕ$)
  \item A data type for formulas
\begin{agda}
data Prop : Set where
  Var  : Fin n → Prop           -- Variables.
  $⊤$    : Prop                   -- Top (truth).
  $⊥$    : Prop                   -- Bottom (falsum).
  _$∧$_  : ($φ$ $ψ$ : Prop) → Prop    -- Conjunction.
  _$∨$_  : ($φ$ $ψ$ : Prop) → Prop    -- Disjunction.
  _$⇒$_  : ($φ$ $ψ$ : Prop) → Prop    -- Implication.
  _$⇔$_  : ($φ$ $ψ$ : Prop) → Prop    -- Biimplication.
  $¬$_   : ($φ$ : Prop) → Prop      -- Negation.
\end{agda}
\end{itemize}
\end{frame}

\begin{frame}[fragile, label=agda-prop]{\texttt{Agda-Prop} Library
  \footnote{\url{https://github.com/jonaprieto/agda-prop}}}
\begin{itemize}
  \item A data type for theorems
\begin{agda}
data _$⊢$_ : ($Γ$ : Ctxt)($φ$ : Prop) → Set
\end{agda}

\item Constructors
\begin{agda}
assume, axiom, weaken, $⊤$-intro, $⊥$-elim, $¬$-intro,
$¬$-elim, $∧$-intro, $∧$-proj$₁$, $∧$-proj$₂$, $∨$-intro$₁$,
$∨$-intro$₂$, $∨$-elim, $⇒$-intro, $⇒$-elim, $⇔$-intro,
$⇔$-elim$₁$, $⇔$-elim$₂$.
\end{agda}
  \item Natural deduction proofs for more than 71 theorems
\begin{agda}
$⇔$-equiv, $⇔$-assoc, $⇔$-comm, $⇒-⇔-¬∨$, $⇔-¬$-to-$¬$,
$¬⇔$-to-$¬$, $¬¬$-equiv, $⇒⇒-⇔-∧⇒$, $⇔$-trans, $∧$-assoc,
$∧$-comm, $∧$-dist, $¬∧$-to-$¬∨¬$, $¬∨¬$-to-$¬∧$, $¬∨¬-⇔-¬∧$,
subst$⊢∧₁$, subst$⊢∧₂$, $∨$-assoc, $∨$-comm, $∨$-dist,
$∨$-equiv, $¬∨$-to-$¬∧¬$, $¬∧¬$-to-$¬∨$, $∨$-dmorgan,
$¬¬∨¬¬$-to-$∨$, cnf, nnf, dnf, RAA, $\cdots$
\end{agda}
\end{itemize}
\end{frame}

\begin{frame}[fragile, label=agda-metis]{\texttt{Agda-Metis} Library
  \footnote{\url{https://github.com/jonaprieto/agda-metis}}}
\vfill
\begin{table}[!ht]
\begin{center}
\scalebox{0.7}{
{\renewcommand{\arraystretch}{1.3}%
\begin{tabular}{|lll|}
\hline
\textbf{Rule} & \textbf{Purpose} &\textbf{Theorem}\\
\hline
\texttt{canonicalize} & transforms formulas to CNF, DNF or NNF &\texttt{atp-canonicalize}\\
\texttt{clausify} & performs clausification &\texttt{atp-clausify}\\
\hyperlink{atp-conjunct}{\texttt{conjunct}} & takes a formula from a conjunction &\hyperlink{atp-conjunct}{\texttt{atp-conjunct}}\\
\texttt{negate} & applies negation to the formula &\texttt{atp-negate}\\
\texttt{resolve} & applies theorems of resolution &\texttt{atp-resolve}\\
\texttt{simplify} & applies over a list of formula to simplify them &\texttt{atp-simplify}\\
\texttt{strip} & splits a formula into subgoals &\texttt{atp-strip}\\[1ex]
\hline
\end{tabular}}
}
\end{center}
\label{tab:agda-metis-table}
\end{table}
\vfill
\end{frame}

\begin{frame}[fragile, label=atp-conjunct]{\texttt{Agda-Metis}: Conjunct Inference
  \footnote{\url{https://github.com/jonaprieto/agda-metis}}}
\begin{itemize}
\item Definition
  \begin{equation*}
  conjunct (ϕ₁ ∧  ϕ₂ ∧ ⋯ ∧ \mathbf{ϕᵢ} ∧ ⋯ ∧ \phi_n, \mathbf{ϕᵢ}) ⟶ \mathbf{ϕᵢ}
  \end{equation*}

\item Function\footnote{\label{note-conjunct}Excerpt from the \texttt{Agda-Metis}
library available in  \texttt{ATP.Metis.Rules.Conjunct} module}:
\begin{agda}
conjunct : Prop → Prop → Prop
conjunct $(φ \wedge ψ)$ $ω$ with $⌊$ eq $φ$ $ω$ $⌋$ | $⌊$ eq ψ $ω$ $⌋$
... | true  | _     = $φ$
... | false | true  = $ψ$
... | false | false = conjunct $φ$ $ω$
conjunct $φ$ $ω$       = $φ$
\end{agda}
\item Theorem\textsuperscript{\ref{note-conjunct}}
\begin{agda}
atp-conjunct
  : ∀ $\{Γ\}$ $\{φ\}$
  → $(ω : Prop)$
  → $Γ \vdash φ$
  → $Γ \vdash$ conjunct $φ$ $ω$
\end{agda}
\end{itemize}
\end{frame}

@@

atp-conjunct $\{Γ\}$ $\{φ ∧ ψ\}$ $ω$ $Γ⊢φ$ with $⌊$ eq $φ$ $ω$ $⌋$ | $⌊$ eq $ψ$ $ω$ $⌋$
... | true  | _     = $∧$-proj$₁$ $Γ⊢φ$
... | false | true  = $∧$-proj$₂$ $Γ⊢φ$
... | false | false =
  atp-conjunct $\{Γ = Γ\}$ $\{φ = φ\}$ $ω$ ($∧$-proj$₁$ $Γ⊢φ$)
atp-conjunct {_} {Var x}    _ = id
atp-conjunct {_} {$⊤$}      _ = id
atp-conjunct {_} {$⊥$}      _ = id
atp-conjunct {_} {$φ ∨ φ₁$} _ = id
atp-conjunct {_} {$φ ⇒ φ₁$} _ = id
atp-conjunct {_} {$φ ⇔ φ₁$} _ = id
atp-conjunct {_} {$¬\,φ$}   _ = id
\end{agda}
\vfill
\gobackreconstruction
\end{frame}

\begin{frame}[fragile, label=agda-file]{\texttt{Agda} Code Example}
  {Generated by \texttt{Athena} Tool}

\begin{itemize}
  \item The problem is $p ∧ q ⊢ q ∧ p$
  \item In \texttt{TPTP} format
\begin{tptp}
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
\end{tptp}
  \item A natural deduction proof
\begin{columns}
\begin{column}{0.8\textwidth}
 \begin{prooftree}
  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{$∧$-\texttt{proj}$₁$}
  \UnaryInfC{$ϕ$}

  \AxiomC{$ϕ ∧ ψ$}
  \RightLabel{$∧$-\texttt{proj}$₂$}
  \UnaryInfC{$ψ$}
  \RightLabel{$∧$-\texttt{intro}}
  \BinaryInfC{$ψ ∧ ϕ$}
  \end{prooftree}
\end{column}
\begin{column}{0.2\textwidth}  %%<--- here
   \begin{tikzpicture}[overlay, remember picture, scale=2]
% \tikzset{shift={(current page.north)}, xshift=cm,yshift=-2cm}
\node[vertex, fill=black] (d1) {};
\node[vertex, fill=black, right=0.5cm of d1 ] (d2) {};
\node[vertex, fill=black, below=0.2cm of d1] (d3) {};
\node[vertex, fill=black, below=0.2cm of d2] (d4) {};
\node[vertex, fill=black, below right= 0.2cm and  0.25cm of d3] (d5) {};
\draw[-, thick] (d1) to (d3);
\draw[-, thick] (d2) to (d4);
\draw[-, thick] (d3) to (d5);
\draw[-, thick] (d4) to (d5);
\end{tikzpicture}
\end{column}
\end{columns}

@@

\begin{frame}[fragile, allowframebreaks]{\texttt{TSTP} Derivation}
\begin{tptp}
fof(a, axiom, p & q).
fof(goal, conjecture, q & p).
fof(subgoal_0, plain, q,
    inference(strip, [], [goal])).
fof(subgoal_1, plain, q => p,
    inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ q,
inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, (~ q),
    inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p & q,
    inference(canonicalize, [], [a])).
fof(normalize_0_2, plain, q,
    inference(conjunct, [], [normalize_0_1])).
fof(normalize_0_3, plain, $false,
    inference(simplify, [],
        [normalize_0_0, normalize_0_2])).
cnf(refute_0_0, plain, $false,
    inference(canonicalize, [], [normalize_0_3])).
fof(negate_1_0, plain, ~ (q => p),
    inference(negate, [], [subgoal_1])).
fof(normalize_1_0, plain, ~ p & q,
    inference(canonicalize, [], [negate_1_0])).
fof(normalize_1_1, plain, p & q,
     inference(canonicalize, [], [a])).
fof(normalize_1_2, plain, p,
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_3, plain, q,
    inference(conjunct, [], [normalize_1_1])).
fof(normalize_1_4, plain, $false,
    inference(simplify, [],
      [normalize_1_0, normalize_1_2, normalize_1_3])).
cnf(refute_1_0, plain, ($false),
    inference(canonicalize, [], [normalize_1_4])).
\end{tptp}
\end{frame}

\begin{frame}[fragile, label=verified-example]{$p ∧ q ⊢ q ∧ p$}{Reconstructed Proof}
\vfill
Definitions
\begin{agda}
p, q, a, goal, subgoal$₀$, subgoal$₁$ : Prop

-- Axiom.
a = (p $∧$ q)

-- Premise.
$Γ$ : Ctxt
$Γ$ = [ a ]

-- Conjecture.
goal = (q $∧$ p)

-- Subgoals.

subgoal$₀$ = q

subgoal$₁$ = (q $⇒$ p)
\end{agda}
\end{frame}
